---
title: "Climate Trends, South Lake Tahoe, CA"
author: "Evie Vermeer, Sophia Leiker and Eva Wilson"
date: "4/5/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(janitor)
library(lubridate)
library(tsibble)
library(feasts)
library(plotly)
library(Kendall)
library(kableExtra)
library(broom)

```

Reading in data: 
```{r}
slt_temps <-read.csv(here("data","slt_temps.csv"))

slt_precip <-read.csv(here("data","slt_precip.csv"))
```

Cleaning up the data:
```{r}
# there are some columns in which the data is being characterized as "character" instead of "numeric" and therefore those columns have difficulty in running summary stats on. This must be due to the "M"'s present. Converting these columns to numeric format and adding NAs instead. 

# converting columns to numeric format for temps
slt_temps$june <- as.numeric(slt_temps$june)
slt_temps$july <- as.numeric(slt_temps$july)
slt_temps$august <- as.numeric(slt_temps$august)
slt_temps$december <- as.numeric(slt_temps$december)
slt_temps$january <- as.numeric(slt_temps$january)
slt_temps$february <- as.numeric(slt_temps$february)

# converting columns to numeric format for precip
# NOTE THIS CONVERTS ALL T (TRACE) TO NA AS WELL, WHICH IS TECHNICALLY NOT TRUE BUT LESS THAN 0.01 IS INSIGNIFICANT (WE CAN CLAIM)
slt_precip$june <- as.numeric(slt_precip$june)
slt_precip$july <- as.numeric(slt_precip$july)
slt_precip$august <- as.numeric(slt_precip$august)
slt_precip$december <- as.numeric(slt_precip$december)
slt_precip$january <- as.numeric(slt_precip$january)
slt_precip$february <- as.numeric(slt_precip$february)
slt_precip$may <- as.numeric(slt_precip$may)
slt_precip$march <- as.numeric(slt_precip$march)
slt_precip$april <- as.numeric(slt_precip$april)
slt_precip$september <- as.numeric(slt_precip$september)
slt_precip$october <- as.numeric(slt_precip$october)
```

```{r}
# do we want to drop na? nah...
#precip <- slt_precip %>% 
 # clean_names() %>% 
 #drop_na(january, february, march, april, may, june, july, august, september, october, november, december)
```

Grouping into Seasons
```{r}
# group by year and calculate the mean for the columns that coincide with the NOAA Meteorological Seasons
slt_temps_seasonal <- slt_temps %>%
  mutate(spring = mean(c(march, april, may), na.rm = TRUE)) %>% ungroup() %>% 
  mutate(summer = mean(c(june, july, august), na.rm = TRUE)) %>% ungroup() %>% 
  mutate(fall = mean(c(september, october,november), na.rm = TRUE)) %>% ungroup() %>% 
  mutate(winter = mean(c(december, january, february), na.rm = TRUE)) %>% ungroup()

slt_precip_seasonal <- slt_precip %>%
  mutate(spring = mean(c(march, april, may), na.rm = TRUE)) %>% ungroup() %>% 
  mutate(summer = mean(c(june, july, august), na.rm = TRUE)) %>% ungroup() %>% 
  mutate(fall = mean(c(september, october,november), na.rm = TRUE)) %>% ungroup() %>% 
  mutate(winter = mean(c(december, january, february), na.rm = TRUE)) %>% ungroup()

slt_temps_seasonal
slt_precip_seasonal
```

Calculate summary statistics 
```{r}
# calculate annual summary statistics - precipitation
precip_year <- slt_precip_seasonal %>% 
  summarize(max_precip = max(annual, na.rm = TRUE), 
            min_precip = min(annual, na.rm = TRUE), 
            mean_precip = mean(annual, na.rm = TRUE),
            total_yearly_precip = sum(annual))
precip_year

# calculate annual summary statistics - temperature
temps_year <- slt_temps_seasonal %>% 
  summarize(max_temp = max(annual, na.rm = TRUE), 
            min_temp = min(annual, na.rm = TRUE), 
            mean_temp = mean(annual, na.rm = TRUE))
temps_year

# calculate august statistics 
precip_august<- slt_precip_seasonal %>% 
  summarize(max_precip = max(august, na.rm = TRUE), 
            min_precip = min(august, na.rm = TRUE), 
            mean_precip = mean(august, na.rm = TRUE))
precip_august

temps_august<- slt_temps_seasonal %>% 
  summarize(max_temp = max(august, na.rm = TRUE), 
            min_temp = min(august, na.rm = TRUE), 
            mean_temp = mean(august, na.rm = TRUE))
temps_august
```


Plot annual averages (temperature)
```{r}
# smooth yearly ggplot
ggplot(data = slt_temps_seasonal) +
  geom_smooth(method = "lm", 
              color = "gray25", 
              aes(x = year, 
                  y = annual)) +
  geom_point(aes(x = year, 
               y = annual, 
               color = annual), 
           size = 2) +
  scale_color_gradientn(colors = c("deepskyblue4", "firebrick")) +
  theme_minimal(14) +
  theme(legend.position = "none") +
  labs(x = element_blank(), 
       y = "Average Annual Temperature (F)") +
  geom_text(x = 2017, y = 49, label = "Max (49.4)", color = "firebrick", size = 4) +
  geom_text(x = 2002, y = 42.5, label = "Min (43)", color = "deepskyblue4", size = 4)
```
Plot annual averages (Precipitation)
```{r}
# smooth yearly ggplot
ggplot(data = slt_precip_seasonal) +
  geom_smooth(method = "lm", 
              color = "gray25", 
              aes(x = year, 
                  y = annual)) +
  geom_point(aes(x = year, 
               y = annual, 
               color = annual), 
           size = 2) +
  scale_color_gradientn(colors = c("firebrick", "deepskyblue4")) +
  theme_minimal(14) +
  theme(legend.position = "none") +
  labs(x = element_blank(), 
       y = "Average Annual Precipitation (in)") + 
  geom_text(x = 2017, y = .123, label = "Max (.12)", color = "deepskyblue4", size = 4) +
  geom_text(x = 2012, y = .023, label = "Min (.02)", color = "firebrick", size = 4)
```
Plot august averages??
```{r}
# precipitation

# temperature 
```

### Statistical Testing

**Table 1** displays the results of statistical testing performed on the climate trends data. Slope and P value refer to the results of simple linear regression. M-K significance level refers to the results of Mann-Kendall test performed on each measure.

JUST COPIED LUCAS CODE HERE - BUT THINK WE NEED TO GO BACK AND EDIT MY SUMMARY STATISTICS ABOVE? RIGHT NOW THEY ONLY GENERATE ONE MIN MAX AND AVERAGE VALUE
```{r}
# linear regressions

# max temp
temp_max_lm <- lm(mean_daily_max_temp ~ year, data = seattle_yearly)
temp_max_tidy <- tidy(temp_max_lm)

# min temp
temp_min_lm <- lm(mean_daily_min_temp ~ year, data = seattle_yearly)
temp_min_tidy <- tidy(temp_min_lm)

# annual precip
yearly_precip_lm <- lm(total_yearly_precip ~ year, data = seattle_yearly)
yearly_precip_tidy <- tidy(yearly_precip_lm)

lm_tidy_all <- rbind(temp_max_tidy, 
                     temp_min_tidy, 
                     yearly_precip_tidy,
                     monthly_precip_tidy)
lm_tidy_all_coef <- lm_tidy_all %>% 
  filter(term == "year") %>% 
  select(-statistic, -std.error, -term)
         
names_vector <- c("Average Maximum Daily Temperature (F)", "Average Minimum Daily Temperature (F)", "Annual Precipitation (inches)")

lm_tidy_all_clean <- data.frame(names_vector, lm_tidy_all_coef)

# Mann Kendall testing for significance
mk1 <- MannKendall(seattle_yearly$mean_daily_max_temp)
mkt1 <- mk1$sl
mk2 <- MannKendall(seattle_yearly$mean_daily_min_temp)
mkt2 <- mk2$sl
mk3 <- MannKendall(seattle_yearly$total_yearly_precip)
mkt3 <- mk3$sl
mk4 <- MannKendall(seattle_monthly$monthly_precip)
mkt4 <- mk4$sl

mk_vector <- c(mkt1, mkt2, mkt3, mkt4)

stats_table <- data.frame(lm_tidy_all_clean, mk_vector)

kable(stats_table, 
      col.names = c("Measurement", "Slope", "P value", "M-K Significance Level"), 
      digits = c(0, 4, 4, 4)) %>% 
  kable_minimal(full_width = FALSE)
```
```

